### 2022-06-06 tex/ill.tex
[ writing ] introduction to MALL

@@

\subsection{The $\oc$-modality}

>>>>

While classical linear logic is perhaps better understood, for the sake of this
thesis I focus on intuitionistic linear logic.
The syntactic difference between the two is analogous to the difference between
classical and intuitionistic logic: The latter restricts the former by only
allowing one formula on the right of a sequent.
Classical linear logic is not irrelevant to this thesis, and appears as an
instance of the framework presented in \cref{sec:framework}.
However, the connection of this work to intuitionistic linear logic is more
direct, and reflects the basis of existing semiring-annotated calculi on
(intuitionistic) typed $\lambda$-calculi.

Intuitionistic linear logic is often motivated by its sensitivity to resources.
The fact that, in intuitionistic (and classical) logic, \emph{any} hypothesis
can be discarded or duplicated means that propositions cannot be used to model
things that may be ``possessed'' or ``spent'', despite hypotheses often being
described as things we ``have''.
In contrast, linear logic allows propositions to behave like physical objects,
being moved around but not being created or destroyed (a priori).

Programming applications of linearity include stateful protocols
\citep{Wadler12}, mutually exclusive capabilities (TODO: cite),
and mutation (TODO: cite).
In each of these cases, the act of using up a hypothesis allows us to divide
time into ``before'' and ``after'', with lack of duplication avoiding confusion
over which half we are in, and lack of deletion allowing us to know whether the
corresponding action actually happened.
For example, when modelling mutation, a variable refers to a single state of a
mutable value.
The \texttt{write} operation uses up such a variable, making the old state
inaccessible, and returns a new linear value to represent the new state of the
mutable value.
Such a protocol naturally also supports a \emph{freezing} operation, by which we
relinquish the ability to mutate the value in return for an immutable reference
to the value produced.

@@

\subsection{The multiplicative-additive fragment}

The multiplicative-additive fragment of linear logic (MALL) is the fragment
where all hypotheses are linear (must be used exactly once).
We will extend MALL with the \emph{exponential} modality in
\cref{sec:bang-modality}.
MALL is unable to embed intuitionistic or classical logic (collectively
known as \emph{structural logics}), as MALL is unable to reflect any of the
discarding or duplication that can be done in proofs in structural logics.

In short, the syntax of intuitionistic MALL can be described as intuitionistic
logic with the structural rules of \emph{weakening} and \emph{contraction}
removed.
However, without the presence of weakening and contraction, we have to be more
careful about the rules we state, so as not to accidentally introduce weakening
and contraction admissibly.
The lack of these structural rules also allows us to observe a new phenomenon:
the distinction (at the level of provability) between \emph{additive} and
\emph{multiplicative} formulations of existing connectives (in particular, in
the intuitionistic case, the conjunction connective).

I present MALL in \cref{fig:mall} in a sequent calculus style, as it was
presented by \citet{girard87linear}.

To encode what it means to use a hypothesis \emph{exactly once}, we first need
to decide what counts as a use.
The simplest case is that the identity sequent counts as a single use of its
sole hypothesis, and conversely does not count as a use of any other hypotheses.
For sequential proofs, created by the \TirName{Cut} rule, if we have a proof of
$A$ using $\Gamma$, and a proof of $B$ using $\Delta$ and $A$, then we have a
proof of $B$ transitively using $\Delta$ and $\Gamma$.
The exchange rule Exch says that use is invariant under permutation.

For the logical connectives, we have genuine choices as to what it means to use
them.
Two cases --- disjunction ($\oplus$) and (linear) implication ($\multimap$) ---
are somewhat intuitive from intuitionistic logic.
A canonical proof of a disjunction is a tag and a proof of one of the two
disjuncts.
This suggests that a proof of a disjunction only uses the same hypotheses as
the proof of the disjunct we actually choose, with the other disjunct being
irrelevant.
Correspondingly, when we use a disjunction hypothesis, we will only actually use
one of the cases, so each branch should use the same hypotheses.
For implication, use is sequential like with the \TirName{Cut} rule, and its
left rule is more or less the only choice that allows use of the surrounding
hypotheses.

For conjunction, there are two choices: Either the conjuncts \emph{together} use
all of the hypotheses, or each of the conjuncts \emph{individually} uses all of
the hypotheses.
The former choice gives us the tensor-product ($\otimes$), and the latter choice
gives us the with-product ($\with$).
These products are equivalent up to provability in structural logics but
distinct in linear logic.

\begin{figure}
  \begin{align*}
    A, B, C &\Coloneqq X \mid I \mid A \otimes B \mid A \multimap B
              \mid 0 \mid A \oplus B \mid \top \mid A \with B \\
    \Gamma, \Delta, \Theta &\Coloneqq {\cdot} \mid \Gamma, A
  \end{align*}
  \begin{mathpar}
    \ebrule{%
      \infer0[Id]{A \vdash A}
    }

    \ebrule{%
      \hypo{\Gamma \vdash A}
      \hypo{\Delta, A \vdash B}
      \infer2[Cut]{\Gamma, \Delta \vdash B}
    }

    \ebrule{%
      \hypo{\Gamma, B, A, \Delta \vdash C}
      \infer1[Exch]{\Gamma, A, B, \Delta \vdash C}
    }

    \ebrule{%
      \hypo{\Gamma \vdash C}
      \infer1[$I$-L]{\Gamma, I \vdash C}
    }

    \ebrule{%
      \infer0[$I$-R]{{\cdot} \vdash I}
    }

    \ebrule{%
      \hypo{\Gamma, A, B \vdash C}
      \infer1[$\otimes$-L]{\Gamma, A \otimes B \vdash C}
    }

    \ebrule{%
      \hypo{\Gamma \vdash A}
      \hypo{\Delta \vdash B}
      \infer2[$\otimes$-R]{\Gamma, \Delta \vdash A \otimes B}
    }

    \ebrule{%
      \hypo{\Gamma \vdash A}
      \hypo{\Delta, B \vdash C}
      \infer2[$\multimap$-L]{\Gamma, \Delta, A \multimap B \vdash C}
    }

    \ebrule{%
      \hypo{\Gamma, A \vdash B}
      \infer1[$\multimap$-R]{\Gamma \vdash A \multimap B}
    }

    \ebrule{%
      \infer0[$0$-L]{\Gamma, 0 \vdash C}
    }

    \text{(no $0$-R)}

    \ebrule{%
      \hypo{\Gamma, A \vdash C}
      \hypo{\Gamma, B \vdash C}
      \infer2[$\oplus$-L]{\Gamma, A \oplus B \vdash C}
    }

    \ebrule{%
      \hypo{\Gamma \vdash A_i}
      \infer1[$\oplus$-R$_i$]{\Gamma \vdash A_0 \oplus A_1}
    }

    \text{(no $\top$-L)}

    \ebrule{%
      \infer0[$\top$-R]{\Gamma \vdash \top}
    }

    \ebrule{%
      \hypo{\Gamma, A_i \vdash C}
      \infer1[$\with$-L$_i$]{\Gamma, A_0 \with A_1 \vdash C}
    }

    \ebrule{%
      \hypo{\Gamma \vdash A}
      \hypo{\Gamma \vdash B}
      \infer2[$\with$-R]{\Gamma \vdash A \with B}
    }
  \end{mathpar}
  \caption{Multiplicative-additive fragment of linear logic}
  \label{fig:mall}
\end{figure}

Implication ($\multimap$) and the tensor-product ($\otimes$, $I$) comprise the
\emph{multiplicative} fragment, while disjunction ($\oplus$, $0$) and the
with-product ($\with$, $\top$) comprise the \emph{additive} fragment.
Categorically, the additive fragment corresponds to products and coproducts,
while the multiplicative fragment corresponds to multicategorical tensor
products and closure.

\subsection{The $\oc$-modality}\label{sec:bang-modality}

### 2022-06-06 tex/linear.tex
[ more ] thesis writing
[ writing ] introduction to MALL

@@

\section{Mechanisation survey}

>>>>

Linear logic was introduced by \citet{girard87linear} as a tool of proof theory.
Its distinguishing feature is that hypotheses, by default, must be used exactly
once.
I devote a chapter to introducing linear logic because it is better understood
than any individual or general semiring-annotated calculi, and is thus an
important point of comparison.
Linear logic will be an instance of the semiring-annotated systems we study
later.

@@

\section{Mechanisation survey}
\input{linmech.tex}

### 2022-06-06 tex/simple.tex
[ writing ] introduction to MALL

>>>>

In this chapter, I review and justify the family of approaches usually used to
represent simple type systems inside dependently typed proof assistants.
These approaches were first presented by \citet{AR99}, who showed a way of
representing well \emph{scoped} terms in a language with polymorphic recursion,
and extended the representation to well \emph{typed} terms in a language with
dependent types.
The representation of terms relies on indexing on both a context --- giving the
types of all the free variables --- and a type for the term itself.
A basic operation on terms is \emph{simultaneous substitution}, which replaces
each variable in the context by a term in another context.
This simultaneous substitution operation fits the form of composition in a
multicategory.
Specifically, in the case of a simple intuitionistic type system, this is the
composition of a Cartesian multicategory, which allows all of the terms being
substituted in to share a context.
